Novel lattice Boltzmann method for simulation of strongly shear thinning viscoelastic fluids

### 2024-03-04 NovelLBM.tex
note
derivation expanded

\section{Semi-analytical solution for Poiseuille flows of PTT fluids with solvent viscosity}
\label{app:semianalyticalPoiseuille}
Our derivation is \RK{hier könnte ein Adjektiv stehen} to the one in Oliveira \cite{oliveira1999}. For both Poiseuille flow and shear flow, we assume that the velocity is unidirectional:
\begin{equation}
	\vec{u} = u_x\left(y\right)\hat{e}_x
\end{equation}
\begin{equation}
	\dot{\gamma}\left(y\right) = \frac{\partial u_x}{\partial y}
\end{equation}

With this the constitutive equation~(\ref{eq:polymerStress}) for $\xi=0$ in steady state reduces to

@@

	\exp\left(\frac{\epsilon\lambda}{\eta_\text{p}}\Tr\ma{\tau}\right)\tau_{xx} &= 2\lambda\dot{\gamma}\tau_{xy} \label{eq:PTTDerivation1}\\
	\exp\left(\frac{\epsilon\lambda}{\eta_\text{p}}\Tr\ma{\tau}\right)\tau_{xy} &= \eta_\text{p}\dot{\gamma} \label{eq:PTTDerivation2}

@@

for a planar flow.
In axisymmetric flows $y$ in eqs.~(\ref{eq:PTTDerivation1}, \ref{eq:PTTDerivation2}) needs to be replaced by the radial coordinate $r$. \RK{2D case is also axisymmetric?}\SG{ no factor 2? }\RK{No, factor 2 comes from the NS eq. and thus is only relevant for Poiseuille solution. This replacement is also much more relevant there as here all ys disappear}
With $\Tr\ma{\tau} = \tau_{xx}$.
Eliminating $\tau_{xy}$ yields the following.

@@

This is solved by the Lambert $W$ function as follows.\todo{split 2D and 3D derivation}

@@

With $\tau_{yy} = 0$ this yields the first normal stress difference $N_1$ as follows. The second normal stress difference is proportional to $\xi$ and therefore $0$.

@@

	N_1 = \sigma_{xx} - \sigma_{yy} = \tau_{xx} = \frac{\eta_\text{p}}{2\epsilon\lambda}W_0\left(4\epsilon \mathrm{Wi}^2\right) %\label{eq:shearN}

@@

	\eta &= \frac{\sigma_{xy}}{\dot{\gamma}} \\
	&= \frac{\tau_{xy} + \eta_\text{s}\dot{\gamma}}{\dot{\gamma}} \\

@@

This gives the relevant solutions for the shear flow. For Poiseuille flow, we start with the Cauchy momentum equation (neglecting gravity), which reads as follows.

@@

	\rho\frac{\Dif\vec{u}}{\Dif t} = -\nabla p + \nabla\cdot\sigma

@@

In steady state for this geometry, once integrated along $y$, this reduces to the following.

@@

	\sigma_{xy} = \frac{\partial p}{\partial x}\frac{y}{2^j}

@@

Here $j$ is $0$ for the two-dimensional case and $1$ in the three-dimensional flow. The total stress experienced by the fluid is the one described by the constitutive equation plus the solvent contribution and thus the following holds.

@@

	\tau_{xy} = \partial_xp\frac{r}{2^j} - \eta_\text{s}\dot{\gamma}

@@

Here we switched from y to writing $y$ to $r$ in order to clarify, that this is a radial coordinate being $0$ in the center of the channel for both cases.

@@

	\tau_{xx} = \frac{2\lambda}{\eta_\text{p}}\tau_{xy}^2 =  \frac{2\lambda}{\eta_\text{p}}\left(\partial_xp\frac{r}{2^j} - \eta_\text{s}\dot{\gamma}\right)^2

@@

		b &= -\sqrt{\frac{W_0\left(2a\right)}{2a}}

>>>>

\section{Semi-analytical solution for Poiseuille flows of PTT fluids with solvent viscosity}
\label{app:semianalyticalPoiseuille}
Our derivation is similar to the one in Oliveira \cite{oliveira1999}. The origin is placed in the center of the channel. The $x$-axis points in flow direction.
We solve both the three-dimensional and the two-dimensional case. In three dimensions we pick cylindrical coordinates as follows.
\begin{align}
	x &= x\\
	y &= r\cos\varphi\\
	z &= r\sin\varphi
\end{align}
This unusual choice is to keep the coordinate systems consistent between the two- and three-dimensional cases. For two dimensions the $y$-axis points along the velocity gradient. This follows from the 3D case limiting $\varphi$ to $\varphi\in\left\{0, \pi\right\}$. Consequently, we define $r = \abs{y}$ in the 2D case. Note, that in 2D this leads to the sign being tracked explicitly as follows.
\begin{equation}
	\frac{\partial}{\partial r} =  \sgn\left(y\right)\frac{\partial}{\partial y}
\end{equation}

For both Poiseuille flow and shear flow, symmetry dictates, that the velocity is unidirectional:
\begin{equation}
	\vec{u} = u_x\left(r\right)\hat{e}_x
\end{equation}
We also define the shear rate as follows.
\begin{equation}
\dot{\gamma}\left(r\right) = \frac{\partial u_x}{\partial r} \leq 0
\end{equation}
Note, that with this the shear rate is always negative. Also, $\dot{\gamma}$ is a constant for pure shear flow.

We start with the solutions of the constitutive equation.
\subsection{Shear solutions}
In two dimensions, the constitutive equation~(\ref{eq:polymerStress}) for $\xi=0$ in steady state easily reduces to the following. Note, that $\ma{\tau}$ is symmetric.
\begin{align}
	\exp\left(\frac{\epsilon\lambda}{\eta_\text{p}}\Tr\ma{\tau}\right)\tau_{xx} &= 2\lambda\sgn\left(y\right)\dot{\gamma}\tau_{xy}\\
	\exp\left(\frac{\epsilon\lambda}{\eta_\text{p}}\Tr\ma{\tau}\right)\tau_{xy} &= \eta_\text{p}\sgn\left(y\right)\dot{\gamma}
\end{align}
With $\Tr\ma{\tau} = \tau_{xx}$. Note that all other components aside from $\tau_{xx}$ and $\tau_{xy}$/$\tau_{yx}$ are zero.
We define $\tau_{xr} = \tau_{xy}\sgn\left(y\right) \leq 0$ to get the following.
\begin{align}
	\exp\left(\frac{\epsilon\lambda}{\eta_\text{p}}\Tr\ma{\tau}\right)\tau_{xx} &= 2\lambda\dot{\gamma}\tau_{xr} \label{eq:PTTDerivation1}\\
	\exp\left(\frac{\epsilon\lambda}{\eta_\text{p}}\Tr\ma{\tau}\right)\tau_{xr} &= \eta_\text{p}\dot{\gamma} \label{eq:PTTDerivation2}
\end{align}

The three-dimensional case yields the same formula using significantly more math.
We set $\xi=0$ and start from the velocity gradient.
	\nabla\vec{u} = \frac{\partial u_x}{\partial r}\hat{e}_r\otimes\hat{e}_x
This, together with the general formula for $\ma{\tau}$,
	\ma{\tau} = \tau_{ij}\hat{e}_i\otimes\hat{e}_j
is used to find the non-zero terms of the upper convected derivative in steady state as follows.
	\begin{split}
		\left(\nabla\vec{u}\right)^\mathrm{T}\cdot\ma{\tau} = \frac{\partial u_x}{\partial r}\hat{e}_x\otimes\hat{e}_r&\left(\tau_{xx}\hat{e}_x\otimes\hat{e}_x+\tau_{xr}\hat{e}_x\otimes\hat{e}_r+\tau_{x\varphi}\hat{e}_x\otimes\hat{e}_\varphi\right.\\
		&+\tau_{rx}\hat{e}_r\otimes\hat{e}_x+\tau_{rr}\hat{e}_r\otimes\hat{e}_r+\tau_{r\varphi}\hat{e}_r\otimes\hat{e}_\varphi\\
		&\left.{ }+\tau_{\varphi x}\hat{e}_\varphi\otimes\hat{e}_x+\tau_{\varphi r}\hat{e}_\varphi\otimes\hat{e}_r+\tau_{\varphi\varphi}\hat{e}_\varphi\otimes\hat{e}_\varphi\right)
	\end{split}
With the orthonormality
	\hat{e}_a\otimes\hat{e}_b\hat{e}_c\otimes\hat{e}_d = \hat{e}_a\hat{e}_b^\mathrm{T}\hat{e}_c\hat{e}_d^\mathrm{T} = \hat{e}_a\left(\hat{e}_b\cdot\hat{e}_c\right)\hat{e}_d^\mathrm{T} = \hat{e}_a\left(\delta_{bc}\right)\hat{e}_d^\mathrm{T} = \delta_{bc}\hat{e}_a\otimes\hat{e}_d
we can reduce this to the following.
	\left(\nabla\vec{u}\right)^\mathrm{T}\cdot\ma{\tau} = \frac{\partial u_x}{\partial r}\left(\tau_{rx}\hat{e}_x\otimes\hat{e}_x+\tau_{rr}\hat{e}_x\otimes\hat{e}_r+\tau_{r\varphi}\hat{e}_x\otimes\hat{e}_\varphi\right)
The second non-zero term in the upper convected derivative due to the symmetry of $\ma{\tau}$ is the transpose of this. Together with the strain rate tensor, which is calculated using the velocity gradient from above, the complete constitutive equation reads:
	\begin{split}
		0 &= \frac{\partial u_x}{\partial r}\left[2\tau_{rx}\hat{e}_x\otimes\hat{e}_x+\tau_{rr}\left(\hat{e}_x\otimes\hat{e}_r+\hat{e}_r\otimes\hat{e}_x\right)+\tau_{r\varphi}\left(\hat{e}_x\otimes\hat{e}_\varphi + \hat{e}_\varphi\otimes\hat{e}_x\right)\right]\\
		&- \exp\left(\frac{\epsilon\lambda}{\eta_\text{p}}\Tr\ma{\tau}\right)\frac{\ma{\tau}}{\lambda} + \frac{\eta_\text{p}}{\lambda}\frac{\partial u_x}{\partial r}\left(\hat{e}_r\otimes\hat{e}_x + \hat{e}_x\otimes\hat{e}_r\right)
	\end{split}
We separate this equation by components, meaning if $\hat{e}_i\otimes\hat{e}_j$ does not appear in this equation $\tau_{ij} = 0$.
	0 = \frac{\partial u_x}{\partial r}\left[2\tau_{rx}\hat{e}_x\otimes\hat{e}_x\right]- \exp\left(\frac{\epsilon\lambda}{\eta_\text{p}}\Tr\ma{\tau}\right)\frac{\ma{\tau}}{\lambda} + \frac{\eta_\text{p}}{\lambda}\frac{\partial u_x}{\partial r}\left(\hat{e}_r\otimes\hat{e}_x + \hat{e}_x\otimes\hat{e}_r\right)
The symmetry of $\ma{\tau}$ also holds in cylindrical coordinates, meaning $\tau_{rx} = \tau_{xr}$. We reorder accordingly.
	\exp\left(\frac{\epsilon\lambda}{\eta_\text{p}}\Tr\ma{\tau}\right)\frac{\ma{\tau}}{\lambda} = \dot{\gamma}\left[2\tau_{xr}\hat{e}_x\otimes\hat{e}_x\right] + \frac{\eta_\text{p}}{\lambda}\dot{\gamma}\left(\hat{e}_r\otimes\hat{e}_x + \hat{e}_x\otimes\hat{e}_r\right)
We separate the components once more. Yielding two unique equations.

@@

	\exp\left(\frac{\epsilon\lambda}{\eta_\text{p}}\Tr\ma{\tau}\right)\tau_{xx} &= 2\lambda\dot{\gamma}\tau_{xr} \\
	\exp\left(\frac{\epsilon\lambda}{\eta_\text{p}}\Tr\ma{\tau}\right)\tau_{xr} &= \eta_\text{p}\dot{\gamma}

@@

The trace con be found to be:
	\Tr\ma{\tau} = \tau_{xx} + \tau_{rr} + \tau_{\varphi\varphi} = \tau_{xx}
As stated above this solution has the same form as for the two-dimensional case. Therefore, the following holds for both cases.
We eliminate $\tau_{xr}$ yielding the following.

@@

This is solved by the Lambert $W$ function as follows.

@@

As already mentioned above most components of $\ma{\tau}$ are zero, using specifically

@@

	0 = \begin{cases}
		\tau_{yy} = \sigma_{yy},\quad\text{in 2D} \\
		\tau_{rr} = \sigma_{rr},\quad\text{in 3D}
	\end{cases}
yields the first normal stress difference $N_1$ as follows. The second normal stress difference is proportional to $\xi$ and therefore $0$.
	N_1 = \sigma_{xx} - \begin{cases}
		\sigma_{yy},\quad\text{in 2D} \\
		\sigma_{rr},\quad\text{in 3D}
	\end{cases} = \tau_{xx} = \frac{\eta_\text{p}}{2\epsilon\lambda}W_0\left(4\epsilon\mathrm{Wi}^2\right) %\label{eq:shearN}

@@

	\eta &= \frac{\sigma_{xr}}{\dot{\gamma}} \\
	&= \frac{\tau_{xr} + \eta_\text{s}\dot{\gamma}}{\dot{\gamma}} \\

@@

With $\sigma_{xr} = \sigma_{xy}\sgn\left(y\right) \leq 0$ in the 2D case analogous to $\tau_{xr}$.
This concludes the relevant solutions for shear flow. However, these equations also hold for Poiseuille flow. Next we calculate the velocity profile for Poiseuille flow.
\subsection{Velocity profile}
We start with the Cauchy momentum equation (neglecting gravity), which reads as follows.

@@

	\rho\frac{\Dif\vec{u}}{\Dif t} = -\nabla p + \nabla\cdot\ma{\sigma}

@@

In steady state for this geometry in two dimensions, this reduces to the following.

@@

	\frac{\partial \sigma_{xr}}{\partial r} = \frac{\partial p}{\partial x}

@@

In three dimensions there is a slight difference. We start from the stress.

@@

	\ma{\sigma} =  \ma{\tau} + 2\eta_\text{s}\ma{D} = \ma{\tau} + \eta_\text{s}\left(\hat{e}_{x}\otimes\hat{e}_{r} + \hat{e}_{r}\otimes\hat{e}_{x}\right)

@@

Therefor as seen for $\ma{\tau}$ above, $\ma{\sigma}$ also only has $\sigma_{xx}$, $\sigma_{xr}$, $\sigma_{rx}$ components. Furthermore, from symmetry $x$ and $\varphi$ derivatives of $\ma{\sigma}$ are zero. This reduces the divergence to the following.
	\nabla\cdot\sigma = \frac{\partial \sigma_{xr}}{\partial r}\hat{e}_x + \frac{\sigma_{xr}}{r}\hat{e}_x
Notably this leads to a different reduced Cauchy momentum equation than for the 2D case.
	\frac{\partial \sigma_{xr}}{\partial r} + \frac{\sigma_{xr}}{r} = \frac{\partial p}{\partial x}
However, due to the convenient form of the additional term both cases are solved by the following.
	\sigma_{xr} = \frac{\partial p}{\partial x}\frac{r}{2^j}
Here $j$ is $0$ for the two-dimensional case and $1$ in the three-dimensional flow. From here the solution is independent of the dimensionality again.
The total stress experienced by the fluid is the one described by the constitutive equation plus the solvent contribution, which can also be written as follows.
	\tau_{xr} = \partial_xp\frac{r}{2^j} - \eta_\text{s}\dot{\gamma}

@@

	\tau_{xx} = \frac{2\lambda}{\eta_\text{p}}\tau_{xr}^2 =  \frac{2\lambda}{\eta_\text{p}}\left(\partial_xp\frac{r}{2^j} - \eta_\text{s}\dot{\gamma}\right)^2

@@

		b &= \sqrt{\frac{W_0\left(2a\right)}{2a}}
